Forest canopies are better generators of biodiversity in the tropics

### 2025-07-28 manuscript.Rmd
adjust figure 1 to show all taxa for scatter plot

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.cap = TRUE, warning=FALSE, message=FALSE, error=FALSE)
library(officedown)
library(officer)
library(tidyverse)
library(spatialreg)
library(data.table)
library(flextable)
library(knitr)
library(chunkhooks)

@@

hook_figure_unit("mm")
```

@@

@@

3 Conservation Research Institue, University of Cambridge, Cambridge, UK

@@

@@

@@

For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns from local to global scales [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a]. In 1799, Alexander von Humbolt began an expedition that would ultimately canonize the role of mountains in global biodiversity patterns as generators of species richness, particularly in tropical regions [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019; @rahbekHumboldtsEnigmaWhat2019]. The extraordinarily high biodiversity of mountains and their enhanced capacity to generate species richness in the tropics has been termed "Humbolt's enigma", and can be attributed to topographic and climatic variation [@rahbekHumboldtsEnigmaWhat2019]. The vertical nature of mountains provides more land surface per unit area than flat regions and produces steep climate gradients over short distances that allow rapid rates of species turnover. Yet mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines that truncate the area of land suitable for diverse communities, and greater seasonal temperature variation, which reduces the climatic distinctness of elevational bands relative to tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013; @janzenWhyMountainPasses1967].

Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation, which is increasingly being recognized as critical to our understanding of biogeographic patterns [@macarthurBirdSpeciesDiversity1961; @fredstonReimaginingSpeciesMove2025; @xingEcologicalPatternsProcesses2023a]. Tall, multi-layered forests generate biodiversity through similar mechanisms to mountains. By increasing habitat volume and producing vertical zones of microclimate and resource availability, tall forests allow species to vertically stratify habitat akin to multi-floored apartments [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. Thus, taller forests increase niche opportunities, which may allow additional species to join the assemblage and lead to greater richness within a given planimetric area compared to regions with shorter vertical vegetation structure [@xingEcologicalPatternsProcesses2023a; @macarthurBirdSpeciesDiversity1961; @macarthurLimitingSimilarityConvergence1967]. This vertical stacking of species has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020; @decontoCharacterizingStructuralComplexity2024]. Yet positive impacts of canopy height on richness are not consistent globally, with often weaker relationships in high latitude regions where forests of comparable vertical profiles to those in the tropics harbor fewer species [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020]. The capacity for forests to generate biodiversity via vertical stratification, yet the variation in their capacity to do so globally therefore presents a vertical dimension to Humbolt's enigma.

Despite decades of ecological inquiry into the vertical dimensions of habitat structure and biodiversity [@macarthurBirdSpeciesDiversity1961; @scheffersIncreasingArborealityAltitude2013; @xingEcologicalPatternsProcesses2023a; @rollLinkingVertebrateSpecies2015], the mechanisms that help or hinder the capacity for tall forests to generate biodiversity remain poorly understood yet are critically important for understanding biogeographic patterns, impacts of climate and landuse change on biodiversity, and the conservation value of forests. Following upon explanations to Humbolt's enigma [@rahbekHumboldtsEnigmaWhat2019], we fill this gap by examining how vegetative complexity and climate may play key roles in mediating the relationship between canopy height and species richness by impacting the capacity for tall canopies to support vertically stratified and species rich arboreal assemblages. For example, denser layering of vegetation in tropical forests may provide more niche opportunities, allowing tall canopies to support more biodiversity than canopies of equivalent height in temperate regions [@daviesAdvancesAnimalEcology2014; @oliveiraVerticalStratificationInfluences2019]. Moreover, densely layered vegetation produces vertical microclimate gradients via interactions with macroclimate conditions, solar radiation, and wind speed that are characterized by humid and thermally stable understories to dry and thermally variable canopies that experiences both hot and cold temperature extremes. Physiological constraints imposed on arboreal life strategies by these more extreme conditions, combined with the distinctness of vertical climate bands may impact species turnover from the ground to the canopy and thus the capacity for taller canopies to support increasing richness [@bashamVerticalStratificationCollapses2020; @oliveiraVerticalStratificationInfluences2019]. Yet only recently have we been able to address these hypotheses, as our ability to quantify variation in vegetation structure and microclimates globally and examine their impacts on biogeographic patterns in three-dimensional space has increased with the launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018], the development of microclimate models that can estimate 3D microclimates any where on earth [@macleanMicroclimfFastCanopy2024], and the collation of large ecological databases on vertical habitat use [@mouraPhylogenyinformedCharacterisationGlobal2024a; @wilmanEltonTraitsSpecieslevelForaging2014].

In this study we combine range maps of over 24,000 terrestrial vertebrate species with comprehensive data on vertical habitat use from the TetrapodTraits and EltonTraits databases [@mouraPhylogenyinformedCharacterisationGlobal2024a; @wilmanEltonTraitsSpecieslevelForaging2014] to examine how forest structure and climate may mediate the capacity for tall forests to support species rich assemblages via impacts on the vertical structure of vertebrate communities. We first investigated the capacity for taller canopies to support increasing species richness in forests and less densely vegetated woody biomes across latitudes. We find that tall canopies are better generators of biodiversity in the tropics, particularly in moist forests (Fig. \ref(fig:fig1)). We then examined the processes driving this pattern using a combination of global models of vertical assemblage structure parameterized with macroclimate and forest structure data and process-based microclimate models [@macleanMicroclimfFastCanopy2024; @macleanMicropointCanopyPoint2025] for smaller regions that integrate forest structure data derived from NASA's Global Ecosystem Dynamics Investigation (GEDI) [@dubayahGlobalEcosystemDynamics2020] with high temporal resolution macroclimate data [@hersbachERA5HourlyData2023]. While the large computational requirements of resolving micrometeorological processes within forest canopies inhibit the use of global 3D climate data in our analysis, site-specific models spanning six regions across tropical, temperate, and boreal forests provide insight into the fine-scale processes underlying large-scale patterns.

# Tall canopies generate biodiversity, but are better generators of biodiversity in the tropics

@@

To evaluate the capacity for tall canopies to generate biodiversity, we fit negative binomial regression models of vertebrate richness as a function of canopy height globally for all vertebrate combined and each vertebrate class (birds, mammals, reptiles, and amphibians). We then mapped the deviance residuals of these models to determine where canopy height over- or under-predicted species richness. To examine variation in the capacity for tall canopies to support rich assemblages across latitudes, we produces separate models for tropical, temperate, and boreal regions in forests and more sparsely vegetated habitats, including woodlands, savannas, and shrublands.

Globally, taller trees, which increase habitat volume, support more species (Fig. S\@ref(fig:global-chrich)). However, global models of vertebrate richness by canopy height overpredict richness at high latitudes (negative residuals) and underpredict richness in the tropics (positive residuals) (Figs. \@ref(fig:fig1)c, S\@ref(fig:resids-taxa)). Therefore, canopies of equivalent height generally harbor more species in the tropics than in temperate and boreal regions. Furthermore, increasing canopy height yields greater gains in richness in the tropics than at higher latitudes (Fig. \ref(fig:fig1)d). For example, an increase in canopy height from 15 m to 25 m increases richness by `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Birds"])` birds and `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Amphibians"])` amphibians in tropical forests and by only `r round(ratedifs$dif[ratedifs$biome2 == "Boreal" & ratedifs$taxa=="Birds"])` birds and `r round(ratedifs$dif[ratedifs$biome2 == "Boreal" & ratedifs$taxa=="Amphibians"])` amphibians in boreal regions. Within the tropics, increasing canopy height generally yields greater gains in richness in forests than in drier woodlands, savannas, and shrublands, though baseline richness in areas with short vegetation are similar (Fig. \@ref(fig:fig1)). For the same increase in canopy height, the increase in bird richness in tropical forests exceeds the increase in bird richness in tropical woodlands, savannas, and shrublands by `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Birds"]) - round(ratedifs$dif[ratedifs$biome2 == "Tropical woodland, savanna, and shrubland" & ratedifs$taxa=="Birds"])` species. The difference in the capacity for tall canopies to generate biodiversity between densely and sparsely vegetated tropical regions is particularly evident in Madagascar where the global model underpredicts vertebrate richness in the humid forests of the east coast and overpredicts richness in the drier woodlands and shrublands of the southwest (Figs. \@ref(fig:fig1)c, S\@ref(fig:resids-taxa)). Thus, increases in canopy height have greater positive impacts on richness in the wet tropics and lower impacts in drier or higher latitude regions, mirroring impacts of mountains [@rahbekHumboldtsEnigmaWhat2019]. Simply put, taller canopies generate biodiversity globally, but are more effective generators of biodiversity in the wet tropics.

[ADD MODEL TABLES TO SUPPLEMENT]

@@

The vertical layering of species within forests may mediate the capacity for tall forests to support rich assemblages [@oliveiraVerticalStratificationInfluences2019; @macarthurBirdSpeciesDiversity1961; @xingEcologicalPatternsProcesses2023a]. As forests increase in height, additional arboreal species may be added to the assemblage in novel canopy space [@macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @larueDiversityVolumeRelationships2023]. An increase in the number of arboreal species without an equivalent increase in terrestrial or fossorial species increases the verticality of the assemblage, which represents the skew in the vertical distribution of species toward the canopy [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017]. Within wooded ecosystems, we therefore expect regions exhibiting high verticality to support species rich assemblages and steep increases in richness with canopy height [@oliveiraVerticalStratificationInfluences2019]. To quantify the verticality of an assemblage, we use the aforementioned trait databases, assigning verticality scores to each species based on their occurrence in fossorial (0), terrestrial (0.5), and arboreal (1) habitats [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017]. For species occurring in multiple habitats, we averaged scores across habitats. We then calculated assemblage verticality as the standard effect size of the average verticality score across species in an assemblage, using 100 simulations to determine if assemblages are more vertical than expected by random chance given their richness and the evolutionary history of the biogeographic realm [@oliveiraVerticalStratificationInfluences2019] (Supplementary methods).

Verticality peaked in the wet tropics and declined toward the poles, exhibiting strong correlations with richness across taxa (Fig. \@ref(fig:fig2), S\@ref(fig:vertmean-biome)). Tropical and subtropical moist forests contained assemblages with an average of `r parb.wettropics*100`% of species exhibiting arboreal strategies, corresponding to some of the most species rich regions (Fig. S\@(fig:meanprop-biome)). Though temperate and boreal regions contain tall forests, albeit not quite reaching the heights of the tallest tropical canopies, they contain assemblages with much lower verticality, and higher proportions of species exhibiting terrestrial and fossorial strategies (Fig. \@ref(fig:fig2)). This pattern suggests that the capacity for taller canopies to support higher richness declines when canopies cannot support highly vertical species assemblages. Understanding environmental factors that constrain the evolution of diverse arboreal assemblages is therefore key to unraveling the vertical dimension to Humbolt's enigma.

# Vegetation structure and canopy climates impact the capacity for tall canopies to support vertical communities

To examine the environmental factors influencing assemblage verticality, we used spatial linear mixed models to model assemblage verticality for each vertebrate class as a function of canopy height and its interaction with macroclimatic factors that may inhibit or enhance the capacity for canopies to support diverse arboreal assemblages. Macroclimate variables included those that may induce physiological constraints - minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, and precipitation of the warmest month - as well as diurnal temperature range, which may influence the distinctness of vertical climate banding (Supplementary methods). We additionally included vegetation density in the models, which impacts the availability of habitat volume. While foliage height diversity and temperature and precipitation seasonality may also impact vertical niche opportunities and the distinctness of climate banding, these variables where strongly correlated with canopy height, minimum temperature, and dry season precipitation, respectively, and were therefore excluded from the models. The macroclimate variables included in these models represent climate conditions experienced in open habitats rather than the microclimate conditions across vertical forest profiles [@defrenneWeatherStationsLack2016; @bramerAdvancesMonitoringModelling2018]. While using macroclimate data was necessary due to the lack of global vertically resolved microclimate data, its use inhibits our ability to understand how microclimates experienced by forest-dwelling organisms impact vertical assemblage structure and richness patterns [@klingesProximalMicroclimateMoving2024; @zellwegerForestMicroclimateDynamics2020]. We therefore additionally modeled vertical microclimate profiles in six 1° x 1° grid cells within forested regions from tropical to boreal latitudes. To do so, we integrated information on the vertical distribution of vegetation derived from NASA's GEDI spaceborne lidar mission [@dubayahGlobalEcosystemDynamics2020] with ERA5 macroclimate data at a high temporal resolution [@hersbachERA5HourlyData2023] in a process-based microclimate model that resolves within-canopy micrometeorological processes [@macleanMicroclimfFastCanopy2024; @macleanMicropointCanopyPoint2025]. To reduce variability in vertical profiles not attributed to latitudinal differences, we restricted microclimate models to points below 500 m a.s.l and that represented the top 75th to 98th quantiles of canopy height within the modeled region (Supplementary methods). By estimating the climate conditions experienced by forest-dwelling organisms, these models provide insight into the microclimatic processes that underlie macroclimate patterns and allow canopies to support vertically stratified and species rich assemblages.

Our models of verticality indicate that taller canopies support highly vertical vertebrate assemblages, while dry and thermally variable macroclimate conditions, including extreme hot and cold temperatures limit verticality and/or the capacity for taller forests to support assemblages with higher verticality, albeit to varying degrees across vertebrate classes (Fig. \@ref(fig:fig3), Table S\@ref(tab:vert-model-results)). These patterns arise due to impacts of climate on vegetation structure and physiological constraints to arboreal lifestyles, as well as impacts of temporal climate variation on the distinctness of vertical climate bands.

@@

Climate impacts the structural complexity of vegetation, which plays a key role in the the availability of arboreal niche opportunities and is typically associated with high animal richness [@xingEcologicalPatternsProcesses2023a; @daviesAdvancesAnimalEcology2014]. Our models indicated that high dry season precipitation positively impacts assemblage verticality across taxa (Fig. \@ref(fig:fig3)). Wet climates with low precipitation seasonality enable the development of tall and structurally complex forests with densely layered vegetation and generally high tree diversity [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024; @hakkenbergClimateMediatesRelationship2021]. Wet climates also support diverse epiphyte communities, which provide additional microhabitats in the canopy that contribute to high arboreal richness [@taylorVascularEpiphytesContribute2022; @thomsenSecondaryFoundationSpecies2018; @ellwoodDoublingEstimateInvertebrate2004]. Hotspots of structural complexity in wet regions that coincide with some of the tallest forests, including the Choco-Darien, the Amazon, Borneo, and New Guinea [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024], thus provide opportunities for fine-scale niche partitioning, yielding species rich assemblages [@camargoForestVerticalComplexity2018; @gouveiaForestStructureDrives2014; @oliveiraVerticalStratificationInfluences2019].

@@

[ADDITIONAL CITATIONS] - <https://doi.org/10.1016/j.rse.2006.11.016>
-<https://doi.org/10.1111/1365-2745.13944>

@@

Vertical features in the landscape, including mountains and trees bring distinct climates within close proximity of each other [@rahbekHumboldtsEnigmaWhat2019; @leahyVerticalNicheElevation2021; @vinodThermalSensitivityForest2023]. By absorbing and reflecting solar radiation and reducing vertical air mixing, dense forests produce vertical microclimate gradients that capture a remarkable volume of climatic variation given their geographic footprint [@defrenneForestMicroclimatesClimate2021]. These steep climate gradients can support high rates of turnover from the ground to the canopy that yield vertically stratified and species rich assemblages [@bashamVerticalStratificationCollapses2020; @xingSpeciesTurnoverAnt2022; @oliveiraVerticalStratificationInfluences2019; @camargoForestVerticalComplexity2018]. However, more extreme microclimates in the canopy than the understory may place physiological constraints on arboreal lifestyles depending on background macroclimate conditions, effectively truncating the available volume of forest and limiting the capacity for tall forests to support high richness [@oliveiraVerticalStratificationInfluences2019].

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.cap = TRUE, warning=FALSE, message=FALSE, error=FALSE)
library(officedown)
library(officer)
library(tidyverse)
library(spatialreg)
library(data.table)
library(flextable)
library(knitr)
library(chunkhooks)

@@

hook_figure_unit("mm") # sets figure units to mm
```

@@

::: {custom-style="Author"}

@@

3 Conservation Research Institute, University of Cambridge, Cambridge, UK

@@

:::

@@

::: {custom-style="Normal"}

Three-dimensional landscape structure, characterized by mountains and trees, plays a key role in shaping species richness patterns. Taller forests have been proposed to support high levels of biodiversity by increasing habitat volume and producing vertical zones of microclimate and resource availability that allow species to partition niche opportunities from the ground to the canopy. Yet the capacity for taller forests to support greater biodiversity is inconsistent globally. Here, we examine this pattern and the climatic mechanisms from macro to micro scales may mediate the capacity for tall canopies to increase species richness via impacts on the vertical structure of vertebrate communities. Overall, we find that tall canopies support greater biodiversity globally, but high precipitation allows them to yield more rapid increases in richness in the wet tropics by supporting structurally complex vegetation and reducing physiological constraints imposed by dry canopy microclimates.

:::

@@

::: {custom-style="Normal"}

For centuries, ecologists have been intrigued by the role of vertical landscape features in biogeographic patterns [@vonhumboltEssayGeographyPlants2013; @macarthurBirdSpeciesDiversity1961; @bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a]. In 1799, Alexander von Humboldt began an expedition that would ultimately canonize the role of mountains as generators of species richness, particularly in tropical regions [@vonhumboltEssayGeographyPlants2013; @HumboldtsLegacy2019; @rahbekHumboldtsEnigmaWhat2019]. This pattern—'Humboldt's enigma'—can be attributed to topographic and climatic variation [@rahbekHumboldtsEnigmaWhat2019]. The vertical nature of mountains provides more land surface per unit area than flat regions and brings distinct climates within close proximity of each, which produces rapid rates of species turnover. Yet mountains are better generators of biodiversity in the tropics because high latitude mountains have lower elevation snow lines that truncate the area of land suitable for diverse communities, and greater seasonal temperature variation, which reduces the climatic distinctness of elevational bands relative to tropical mountains [@rahbekHumboldtsEnigmaWhat2019; @vonhumboltEssayGeographyPlants2013; @janzenWhyMountainPasses1967].

Over a century later, Robert MacArthur highlighted the role of another vertical dimension in promoting high biodiversity: the vertical structure of vegetation [@macarthurBirdSpeciesDiversity1961]. Tall, multi-layered forests generate biodiversity through similar mechanisms to mountains. By increasing habitat volume and producing vertical zones of microclimate and resource availability, tall forests allow species to vertically stratify habitat akin to multi-floored apartments [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023a; @macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. This vertical stacking of species with increasing arboreal niche opportunities has been described as a mechanistic link between canopy height and species richness, and is thought to contribute to exceptionally high biodiversity in the tropics where canopies are often tall and structurally complex [@oliveiraVerticalStratificationInfluences2019; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020; @decontoCharacterizingStructuralComplexity2024]. Yet taller canopies are not consistently associated with higher biodiversity across the globe [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020]. The capacity for forests to generate biodiversity via vertical stratification, yet the variation in their capacity to do so globally therefore presents 'MacArthur's enigma'.

Despite decades of ecological inquiry into the vertical dimensions of habitat structure and biodiversity [@macarthurBirdSpeciesDiversity1961; @scheffersIncreasingArborealityAltitude2013; @xingEcologicalPatternsProcesses2023a; @rollLinkingVertebrateSpecies2015], the mechanisms that help or hinder the capacity for tall forests to produce species rich assemblages remain poorly understood yet are critically important for understanding biogeographic patterns, impacts of climate and land-use change on biodiversity, and the conservation value of forests. Following explanations of Humboldt's enigma [@rahbekHumboldtsEnigmaWhat2019], we examine how climate from macro to micro scales mediates the relationship between canopy height and species richness by impacting the capacity for tall canopies to support vertically stratified and species rich arboreal assemblages. For example, climatic conditions that promote structurally complex canopies may provide more niche opportunities, allowing tall canopies to support more biodiversity than structurally simpler canopies of equivalent height [@daviesAdvancesAnimalEcology2014; @oliveiraVerticalStratificationInfluences2019]. Moreover, physiological constraints imposed on arboreal life strategies by dry, hot, or cold canopy microclimates combined with the distinctness of vertical climate bands may impact species turnover from the ground to canopy and thus the capacity for taller canopies to support increasing richness [@bashamVerticalStratificationCollapses2020; @oliveiraVerticalStratificationInfluences2019]. Only recently have we been able to address these hypotheses, as our ability to quantify global variation in 3D vegetation structure, microclimates, and animal community structure has increased with the launch of satellite missions to measure forest height and structural complexity [@popkinSpaceLaserMap2018], the development of microclimate models that can estimate 3D microclimates anywhere on Earth [@macleanMicroclimfFastCanopy2024], and the collation of large ecological databases on vertical habitat use [@mouraPhylogenyinformedCharacterisationGlobal2024a; @wilmanEltonTraitsSpecieslevelForaging2014].

We combine range maps of over 24,000 terrestrial vertebrate species with comprehensive data on vertical habitat use from the TetrapodTraits and EltonTraits databases [@mouraPhylogenyinformedCharacterisationGlobal2024a; @wilmanEltonTraitsSpecieslevelForaging2014] to examine how climate from macro to micro scales may mediate the capacity for tall forests to support diverse assemblages via impacts on the vertical structure of vertebrate communities. We first investigate the capacity for taller canopies to support increasing species richness in woody biomes across latitudes. We then examine the processes driving observed patterns using a combination of global models of vertical assemblage structure parameterized with macroclimate and forest structure data and regional models of vertical microclimate profiles [@macleanMicroclimfFastCanopy2024; @macleanMicropointCanopyPoint2025], which integrate forest structure data from NASA's Global Ecosystem Dynamics Investigation (GEDI) [@dubayahGlobalEcosystemDynamics2020] with high temporal resolution macroclimate data [@hersbachERA5HourlyData2023] to estimate temperature and relative humidity from the ground to the canopy.

:::

# Tall canopies generate biodiversity, but more so in the tropics

@@

::: {custom-style="Normal"}

To evaluate the capacity for tall canopies to generate biodiversity, we fit negative binomial regression models of global vertebrate richness as a function of canopy height for all vertebrates combined and each vertebrate class (birds, mammals, reptiles, and amphibians), mapping the deviance residuals to determine where canopy height over- or under-predicted species richness. Globally, taller canopies support more species (Fig. S\@ref(fig:global-chrich); Table S\@ref(tab:richness-mod-results)), because they increase habitat volume and complexity, which allows fine-scale niche partitioning from the ground to the canopy [@larueDiversityVolumeRelationships2023; @macarthurBirdSpeciesDiversity1961; @xingEcologicalPatternsProcesses2023a]. Yet global models of vertebrate richness by canopy height generally overpredict richness at high latitudes (negative residuals) and underpredict richness in the tropics (positive residuals) (Figs. \@ref(fig:fig1)c, S\@ref(fig:resids-taxa)). Therefore, canopies of equivalent height generally harbor more species in the tropics than in temperate and boreal regions.

Models that allow impacts of canopy height to vary across biomes indicate that increasing canopy height yields greater gains in richness in the wet tropics than at higher latitudes or drier tropical biomes (Fig. \@ref(fig:fig1)d). For example, an increase in canopy height from 15 m to 25 m corresponds to an increase of `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Birds"])` bird and `r round(ratedifs$dif[ratedifs$biome2 == "Tropical forest" & ratedifs$taxa=="Amphibians"])` amphibian species in tropical forests, but only `r round(ratedifs$dif[ratedifs$biome2 == "Boreal" & ratedifs$taxa=="Birds"])` bird and `r round(ratedifs$dif[ratedifs$biome2 == "Boreal" & ratedifs$taxa=="Amphibians"])` amphibian species in boreal regions. The difference in impacts of taller canopies on biodiversity between wet and dry tropical regions is particularly evident in Madagascar, where the global model underpredicts vertebrate richness in the humid forests of the east coast and overpredicts richness in the drier woodlands and shrublands of the southwest (Figs. \@ref(fig:fig1)c, S\@ref(fig:resids-taxa)). Thus, increases in canopy height have greater positive impacts on richness in the wet tropics and lower impacts in drier regions or higher latitudes. Simply put, taller canopies generate biodiversity globally, but are more effective generators of biodiversity in the wet tropics - a pattern mirroring that of mountains [@rahbekHumboldtsEnigmaWhat2019].

:::

@@

::: {custom-style="Normal"}

Verticality—the skew in the vertical distribution of species toward the canopy [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017]—peaked in the wet tropics and declined toward the poles, exhibiting strong correlations with richness across taxa (Fig. \@ref(fig:fig2), S\@ref(fig:meanvert-biome)). Tropical and subtropical moist forests contained assemblages with an average of `r parb.wettropics*100`% of species exhibiting arboreal strategies, corresponding to some of the most species rich regions (Fig. S\@ref(fig:meanprop-biome)). Though temperate and boreal regions contain tall forests, albeit not quite reaching the heights of the tallest tropical canopies, they contain assemblages with much lower verticality, and higher proportions of species exhibiting terrestrial and fossorial strategies (Fig. \@ref(fig:fig2)). This pattern suggests that the capacity for taller canopies to support higher richness declines when canopies cannot support highly vertical species assemblages. Understanding environmental factors that constrain the evolution of diverse arboreal assemblages is therefore key to unraveling MacArthur's enigma.

The vertical layering of species within forests may mediate the capacity for tall forests to support diverse assemblages [@oliveiraVerticalStratificationInfluences2019; @macarthurBirdSpeciesDiversity1961; @xingEcologicalPatternsProcesses2023a]. As forests increase in height, additional arboreal species may be added to the assemblage in novel canopy space [@macarthurBirdSpeciesDiversity1961; @rollLinkingVertebrateSpecies2015; @larueDiversityVolumeRelationships2023]. An increase in the number of arboreal species without an equivalent increase in terrestrial or fossorial species increases the verticality of the assemblage, which represents the skew in the vertical distribution of species toward the canopy [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017]. Within wooded ecosystems, we therefore expect regions exhibiting high verticality to support diverse assemblages and steep increases in richness with canopy height [@oliveiraVerticalStratificationInfluences2019]. To quantify the verticality of an assemblage, we used the aforementioned trait databases, assigning verticality scores to each species based on their occurrence in fossorial (0), terrestrial (0.5), and arboreal (1) habitats [@oliveiraVerticalStratificationInfluences2019; @scheffersVerticalArborealityHorizontal2017]. For species occurring in multiple habitats, we averaged scores across habitats. We then calculated assemblage verticality as the standard effect size of the average verticality score across species in an assemblage, using 100 simulations to determine if assemblages are more vertical than expected by random chance given their richness and the evolutionary history of the biogeographic realm [@oliveiraVerticalStratificationInfluences2019] (Supplementary methods).

:::

# Vegetation structure and canopy climates impact the capacity for tall canopies to support vertical assemblages

::: {custom-style="Normal"}

The microclimate models integrate forest structure data derived  While the large computational requirements of resolving micrometeorological processes within forest canopies inhibit the use of global 3D climate data in our analysis, site-specific models spanning six regions across tropical, temperate, and boreal forests provide insight into the fine-scale processes underlying large-scale patterns.

To examine environmental factors influencing assemblage verticality, we used spatial linear mixed models to model assemblage verticality for each vertebrate class as a function of canopy height and its interaction with macroclimatic factors that may inhibit or enhance the capacity for canopies to support diverse arboreal assemblages. Macroclimate variables included those that may impact vegetation structural complexity or place physiological constraints on arboreal life strategies - minimum temperature of the coldest month, maximum temperature of the warmest month, precipitation of the driest month, and precipitation of the warmest month - as well as diurnal temperature range, which may influence the distinctness of vertical climate banding (Supplementary methods). We additionally included vegetation density in the models, which impacts habitat volume. While foliage height diversity and temperature and precipitation seasonality may also impact vertical niche opportunities and the distinctness of climate banding, these variables were strongly correlated with canopy height, minimum temperature, and dry season precipitation, respectively, and were therefore excluded from the models.

While using macroclimate data was necessary due to the lack of global vertically resolved microclimate data, macroclimate maps do not represent the microclimatic conditions experienced across vertical forest profiles that directly influence the vertical structure of animal assemblages [@defrenneWeatherStationsLack2016; @bramerAdvancesMonitoringModelling2018; @scheffersIncreasingArborealityAltitude2013]. We therefore additionally modeled vertical microclimate profiles in six 1° x 1° regions from tropical to boreal forests. To do so, we integrated information on the vertical distribution of vegetation derived from the GEDI spaceborne lidar [@dubayahGlobalEcosystemDynamics2020] with ERA5 macroclimate data at a high temporal resolution [@hersbachERA5HourlyData2023] in a process-based microclimate model that resolves within-canopy micrometeorological processes [@macleanMicroclimfFastCanopy2024; @macleanMicropointCanopyPoint2025] (Supplementary methods). By estimating the climate conditions experienced by forest-dwelling organisms, these models provide insight into the microclimatic processes that underlie macroclimate patterns and allow canopies to support vertically stratified and diverse animal assemblages.

Overall, our models of verticality indicate that taller canopies support highly vertical vertebrate assemblages, while dry and thermally variable macroclimate conditions, including extreme hot and cold temperatures limit verticality and/or the capacity for taller forests to support assemblages with higher verticality, albeit to varying degrees across vertebrate classes (Fig. \@ref(fig:fig3), S\@ref(fig:preds-univariate)-\@ref(fig:preds-interaction); Table S\@ref(tab:vert-model-results)). These patterns arise due to impacts of climate on vegetation structure and physiological constraints to arboreal strategies, as well as impacts of temporal climate variation on the distinctness of vertical climate bands.

:::

@@

::: {custom-style="Normal"}

Climate impacts the structural complexity of vegetation, which influences the availability of arboreal niche opportunities and is typically positively associated with high animal richness [@xingEcologicalPatternsProcesses2023a; @daviesAdvancesAnimalEcology2014; @goetzLaserRemoteSensing2007]. Our models indicated that high dry season precipitation positively affects assemblage verticality across taxa (Fig. \@ref(fig:fig3)). Wet climates with low precipitation seasonality support tall and structurally complex forests with densely layered vegetation [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024], as well as diverse epiphyte communities, which provide additional microhabitats in the canopy that contribute to high arboreal richness [@taylorVascularEpiphytesContribute2022; @thomsenSecondaryFoundationSpecies2018; @ellwoodDoublingEstimateInvertebrate2004]. Hotspots of structural complexity in wet regions that coincide with some of the tallest forests, including the Choco-Darien, the Amazon, Borneo, and New Guinea [@ehbrechtGlobalPatternsClimatic2021a; @decontoCharacterizingStructuralComplexity2024], thus provide opportunities for fine-scale niche partitioning, yielding species rich assemblages [@camargoForestVerticalComplexity2018; @gouveiaForestStructureDrives2014; @oliveiraVerticalStratificationInfluences2019].

@@

:::

@@

::: {custom-style="Normal"}

Vertical features in the landscape, including mountains and trees bring distinct climates within close proximity of each other [@rahbekHumboldtsEnigmaWhat2019; @leahyVerticalNicheElevation2021; @vinodThermalSensitivityForest2023]. By absorbing and reflecting solar radiation and reducing vertical air mixing, dense forests produce vertical microclimate gradients that capture a substantial volume of climatic variation from the cool and wet understory to the climatically variable canopy, which experiences drier, hotter, and colder conditions [@defrenneForestMicroclimatesClimate2021]. These steep climate gradients can support high rates of species turnover from the ground to the canopy that yield vertically stratified, diverse assemblages [@bashamVerticalStratificationCollapses2020; @xingSpeciesTurnoverAnt2022; @oliveiraVerticalStratificationInfluences2019; @camargoForestVerticalComplexity2018]. However, more extreme microclimates in the canopy than the understory may place physiological constraints on arboreal lifestyles depending on background macroclimate conditions, effectively truncating the available volume of forest and limiting the capacity for tall forests to support high richness [@oliveiraVerticalStratificationInfluences2019].
